\thetitle

### 2021-05-20 writing/conclusion.tex
Rephrase start of conclusion section
the rest, minus the revisions of the results and discussion sections

The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
prefer the run-time checking of Natural over other soundness methods. But just
because a random set of programmers expresses this view does not mean that blame
assignment adds value. Similarly, researchers and language designers have
implicitly answered this question one way or another without evidence for the
blame-strategy dimension. The experiment presented in this paper provides some
justification for the programmers' leanings and helps language creators revisit
their decisions. Of course, the design choice remains a trade-off, and the
the experiment contribute to only one of its many facets.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\citep{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create typed facade modules that import untyped
functions and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can easily
type-check typed modules. The problem is that these retroactive additions of
types to a library may result from a misunderstanding of the code. \emph{Any
retroactively ascribed type may thus be a mistake itself.}

The cited evidence suggests that this scenario is quite common and largely
unadressed.  A future evaluation must develop mutators that produce incorrect
type annotations without breaking the code itself. Some preliminary work
suggests that such type mutators are even more difficult to develop than the
type-level code mutators presented here. Based on the gradual typing literature,
the Natural semantics should usually discover such mistakes. In contrast, the
Transient and Erasure semantics cannot help with such mistakes at all; indeed,
we expect that these latter two raise misleading exceptions or produce
plainly-wrong answers.  Of course, only additional experimental work can confirm
or reject these conjectures.

>>>>

The work of~\citet{tgpk-dls-2018} suggests that, by intuition, programmers
prefer the run-time checking of Natural over other soundness methods. But of course,
the opinion of a random set of programmers does not mean that blame
assignment adds value. Similarly, researchers and language designers have
implicitly answered this question one way or another without evidence for the
blame-strategy dimension. The experiment presented in this paper provides some
justification for the programmers' leanings and helps language creators revisit
their decisions. Of course, the design choice remains a trade-off
along several dimensions while the presented experiment sheds light on
only one of them.

The paper does {\em not\/} address a problem in the gradually typed world that
was pointed out early on by practical researchers~\citep{incorrect-ts,
sta-nt-base-types, wmwz-ecoop-2017} and that has recently received theoretical
attention~\citep{gfd-oopsla-2019, cc-oopsla-20}: mistakes in type annotations
themselves.  Developers use gradual typing to move an untyped code base into the
typed realm, and to this end, they need typed APIs for the vast repositories of
already-existing libraries. Instead of converting the libraries themselves,
language implementors merely create facade modules that import untyped functions
and export them with type annotations, like {\tt typed-pack-lib} in
figure~\ref{fig:example}.  With those facades, the compiler can 
type-check typed modules. But, these retroactive additions of types to a library
may result from a misunderstanding of the code. \emph{Any retroactively ascribed
type may thus be a mistake itself.}

The cited evidence suggests that this scenario is quite common and largely
unadressed.  A future evaluation must develop mutators that produce incorrect
type annotations without breaking the code itself. Some preliminary work
suggests that such type mutators are even more difficult to develop than the
type-level code mutators presented here. Based on the gradual typing literature,
the Natural semantics should discover such mistakes. In contrast, the Transient
and Erasure semantics cannot help with such mistakes at all; indeed, we expect
that these latter two raise misleading exceptions or produce wrong answers
without warning.  Only additional experimental work can confirm or reject these
conjectures.

### 2021-05-20 writing/discussion.tex
the rest, minus the revisions of the results and discussion sections

%% -----------------------------------------------------------------------------
% \subsection{Contextualizing the results through an example scenario}

@@

Interpreting the numeric summaries and aggregations of the last section demands
an intuitive understanding of what blame trails look like in practice.  A
concrete example of blame trails and programmer modes is a good basis for
synthesizing this kind of intuition.

@@

In short, this figure concretely demonstrates how the rational programmer
behaves in different modes. In this case, the behaviors differ from each other
in five of the six modes (the two Transient blame modes behave the same).
The reader may keep these scenarios in mind as we next
discuss the numeric results of the last section.

>>>>

%% -----------------------------------------------------------------------------
Interpreting the numeric summaries and aggregations of the preceding section
demands an intuitive understanding of what blame trails look like in practice.
A concrete example of blame trails and programmer modes is a good basis for
synthesizing this kind of intuition.

@@

@@

In short, this figure concretely demonstrates how the rational programmer
behaves in different modes. In this case, the behaviors differ from each other
in five of the six modes (the two Transient blame modes behave the same).
The reader may keep these scenarios for the following discussion of the numeric results.

### 2021-05-20 writing/related.tex
the rest, minus the revisions of the results and discussion sections

\citet{lksfd-popl-2020}'s human-out-of-the-loop empirical analysis of blame for
higher-order contracts is the inspiration for this paper.  While they do not
spell out the notion of the rational programmer, they present the basic
ingredients of the technique.

At a technical level, this work and~\citet{lksfd-popl-2020}'s differ in
many ways. The experiment presented here concerns three different modes of
gradual typing, theirs a single notion of contract checking.  It also
contributes the idea of creating three comparable rational programmers
(with several modes). Thus the method of this paper answers whether blame
adds value to a gradual type system and what gradual type system it
benefits the most; while~\citet{lksfd-popl-2020} only establish a
systematic relation between blame for contracts and behavioral bugs. The
type-level mutators represent another technical contribution of the work
presented here; as explained in section~\ref{sec:mutate}, almost none
of~\citet{lksfd-popl-2020}'s mutators is useful at the level of type
mistakes.

The literature on gradual typing presents many semantics beyond the three used
here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
 Monotonic References~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differ from Natural in the treatment of mutable data by
associating heap locations with a type and permitting only updates that 
make that type more precise.  
Among these representative semantics, only Amnesic, Nom and Monotonic References present innovative blame
strategies.  The experiment excludes the first because it is
merely a theoretical construction, the second because it imposes severe restrictions on
programmers and the third because it would require a complete
re-engineering of the Racket runtime. That said, the paper offers a
detailed chart that the creators of these blame strategies can transplant to their
linguistic setting and determine the value of their blame strategies.

>>>>

At a philosophical level, \citet{lksfd-popl-2020} present the first empirical
analysis {\em without\/} involving humans in this general area.  While they do
not spell out the notion of the rational programmer, they present many of the
basic ingredients.

At a technical level, the two pieces of work differ in many ways.  First the
experiment presented here involves three different modes of gradual typing,
theirs a single notion of contract checking.  Second, this paper also
contributes the idea of creating three comparable rational programmers, with
several modes. Hence it can answer whether blame adds value to a gradual type
system and which gradual type system it benefits the most, while theirs
establishes only a systematic relation between blame for contracts and
behavioral bugs. Finally, this paper also contributes type-level mutators. As
explained in section~\ref{sec:mutate}, almost none of~\citet{lksfd-popl-2020}'s
mutators is useful in the context of gradual typing.

The literature on gradual typing presents many semantics beyond the three used
here and three additional blame strategies.  Pyret (pyret.org) assigns fixed-size
data types the Natural semantics and functions a Transient semantics. The
Forgetful~\citep{cl-icfp-2017} and the Amnesic~\citep{gfd-oopsla-2019} semantics
are similar to Transient but use wrappers instead of in-lined checks.
Nom~\citep{mt-oopsla-2017} and other \emph{concrete\/}
semantics~\citep{wnlov-popl-2010, rsfbv-popl-2015, rzv-ecoop-2015,
rat-oopsla-2017} assume that every value comes with a type tag and use tag
checks to supervise the interactions between typed and untyped code.  The semantics~\citep{tlt-popl-2019,
etg-icfp-19, tt-scp-20, tgt-popl-18, tt-sas-17} derived with the Abstracting
Gradual Typing technique~\citep{gct-popl-2016} are variants of Natural.
The Monotonic semantics~\citep{svctg-esop-2015, rsfbv-popl-2015, sfrbcsb-popl-2014, kas-pldi-2019} 
differs from Natural in the treatment of mutable data. It 
associates every heap location with a type and permits only updates that 
do not make the type less precise.  
Among these semantics, only Amnesic, Nom, and Monotonic present interesting blame
strategies.  The experiment excludes the first because it is 
merely a theoretical construction, the second because it imposes severe restrictions on
programmers, and the third because it requires a re-engineering of the Racket runtime.

### 2021-05-20 writing/sample.tex
L says it's a typo
the rest, minus the revisions of the results and discussion sections

To become computationally feasible and statistically sound, the experiment employs
stratified uniform random sampling with three levels of stratification.
The first two levels group mutants first by benchmark and then by
mutator, making sure that the sample within each benchmark reflects the diversity of
mutants with respect to the mutators that generated them.  Specifically, the
experiment samples 80 interesting mutants per benchmark,
evenly-distributed across all of the mutators that contribute mutants for the
benchmark. Some benchmarks have less than  80 mutants with interesting scenarios,
in which case the only choice is to include them all.
The result is a total of 756 interesting mutants across all benchmarks.
Finally, the third level of sampling draws 96 random interesting debugging
scenarios from each configuration lattice with replacement. The
final sample thus consists of 72,192 interesting scenarios.

%% MF: I don't get this number ^^^^
% > (* 756 96)
% 72576

>>>>

To render the experiment computationally feasible and statistically sound, it
becomes necessary to sample this large space in a uniform and stratified manner.
The first two levels of stratification group mutants first by benchmark and then
by mutator, making sure that the sample within each benchmark reflects the
diversity of mutants with respect to the mutators that generated them.
Specifically, the experiment samples 80 interesting mutants per benchmark,
evenly-distributed across all of the mutators that contribute mutants for the
benchmark. Some benchmarks have less than 80 mutants with interesting scenarios,
in which case the only choice is to include them all.  The result is a total of
752 interesting mutants across all benchmarks.  Finally, the third level of
sampling randomly draws 96 debugging scenarios from each configuration lattice
\mf{is ``with replacement'' correct here? Do you put it back for the next draw?}
with replacement. The final sample thus consists of 72,192 interesting
scenarios.
